---
title: "Interactive Final SuperLearner Event Prediction Report: Type Z"
author: "Event ENM workflow"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 11,
  fig.height = 3.75,
  dpi = 150,
  out.width = "100%"
)

# Help R find packages installed in the per-user Windows library, e.g. R/win-library/4.6.
LOCALAPPDATA_DIR <- normalizePath(Sys.getenv("LOCALAPPDATA"), winslash = "/", mustWork = FALSE)
WINDOWS_USER_R_LIB <- file.path(
  LOCALAPPDATA_DIR,
  "R",
  "win-library",
  paste(R.version$major, strsplit(R.version$minor, "\\.")[[1]][1], sep = ".")
)
if (dir.exists(WINDOWS_USER_R_LIB) && !WINDOWS_USER_R_LIB %in% .libPaths()) {
  .libPaths(c(WINDOWS_USER_R_LIB, .libPaths()))
}

required_packages <- c("leaflet", "terra", "htmltools", "htmlwidgets")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop("Install required packages before knitting this report: ", paste(missing_packages, collapse = ", "), call. = FALSE)
}

library(leaflet)
library(htmltools)

get_current_script_path <- function() {
  command_args <- commandArgs(trailingOnly = FALSE)
  file_arg <- grep("^--file=", command_args, value = TRUE)
  if (length(file_arg) > 0) {
    return(normalizePath(sub("^--file=", "", file_arg[1]), winslash = "/", mustWork = TRUE))
  }
  for (frame in rev(sys.frames())) {
    if (!is.null(frame$ofile)) {
      return(normalizePath(frame$ofile, winslash = "/", mustWork = TRUE))
    }
  }
  if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
    active_doc <- tryCatch(rstudioapi::getActiveDocumentContext()$path, error = function(e) NA_character_)
    if (!is.na(active_doc) && nzchar(active_doc)) {
      return(normalizePath(active_doc, winslash = "/", mustWork = TRUE))
    }
  }
  NA_character_
}

find_code_dir <- function() {
  script_path <- get_current_script_path()
  wd <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  script_dir <- if (!is.na(script_path)) dirname(script_path) else NA_character_
  wd_parents <- normalizePath(
    file.path(wd, c(".", "..", "../..", "../../..")),
    winslash = "/",
    mustWork = FALSE
  )
  script_candidates <- if (!is.na(script_dir)) {
    normalizePath(file.path(script_dir, c(".", "..", "../..")), winslash = "/", mustWork = FALSE)
  } else {
    character(0)
  }
  candidates <- unique(normalizePath(c(script_candidates, wd_parents, file.path(wd_parents, "KSPH Code")), winslash = "/", mustWork = FALSE))
  for (candidate in candidates) {
    if (
      file.exists(file.path(candidate, "config", "predictor_list.csv")) &&
      dir.exists(file.path(candidate, "R_python_code"))
    ) {
      return(candidate)
    }
  }
  stop(
    "Could not locate the KSPH Code repo root. Knit this report from inside the KSPH Code project, ",
    "or set CODE_DIR manually near the top of 09Z_interactive_prediction_report_email.Rmd.",
    call. = FALSE
  )
}

sanitize_path_component <- function(value) {
  value <- trimws(as.character(value))
  value <- gsub("[^A-Za-z0-9]+", "_", value)
  value <- gsub("^_+|_+$", "", value)
  if (!nzchar(value)) {
    stop("Analysis folder name cannot be blank after sanitizing.", call. = FALSE)
  }
  value
}

existing_path <- function(paths) {
  normalized <- normalizePath(paths, winslash = "/", mustWork = FALSE)
  found <- normalized[file.exists(normalized)]
  if (length(found) == 0) {
    return(NA_character_)
  }
  found[1]
}

CODE_DIR <- find_code_dir()

STUDY_AREA_ANALYSIS_NAME <- "equatorial_africa"
FINAL_ANALYSIS_GROUP <- "type_Z"
FINAL_MODEL_FAMILY_NAME <- "final_sl"
MAP_SET_NAME <- "maps"

ACTIVE_STUDY_AREA_ANALYSIS_NAME <- sanitize_path_component(STUDY_AREA_ANALYSIS_NAME)
FINAL_ANALYSIS_GROUP <- sanitize_path_component(FINAL_ANALYSIS_GROUP)
ANALYSIS_DIR <- file.path(CODE_DIR, "analyses", ACTIVE_STUDY_AREA_ANALYSIS_NAME)
DATA_DIR <- file.path(ANALYSIS_DIR, "data")
ANALYSIS_OUTPUT_DIR <- file.path(ANALYSIS_DIR, "outputs")
PREDICTION_DIR <- file.path(
  ANALYSIS_OUTPUT_DIR,
  FINAL_MODEL_FAMILY_NAME,
  FINAL_ANALYSIS_GROUP,
  MAP_SET_NAME,
  "pred"
)

discover_raster_years <- function(subdir, pattern, replacement = "\\1") {
  paths <- list.files(file.path(PREDICTION_DIR, subdir), pattern = pattern, full.names = FALSE)
  years <- suppressWarnings(as.integer(sub(pattern, replacement, paths)))
  sort(unique(years[is.finite(years)]))
}

PREDICTION_YEARS <- discover_raster_years("ror", "^ror_([0-9]{4})\\.tif$")
ANOMALY_YEARS <- discover_raster_years("hist_anom", "^hist_anom_([0-9]{4})\\.tif$")
MAP_OPACITY_VALUE <- 0.72
MAP_OPACITY_TOP1 <- 0.92
LEAFLET_MAP_HEIGHT <- "300px"
LEAFLET_NA_COLOR <- "#00000000"

OUTCOMES_CSV_CANDIDATES <- c(
  file.path(DATA_DIR, "outcomes_csv.csv"),
  file.path(CODE_DIR, "config", "outcomes_csv.csv")
)
EVENT_OVERLAY_GROUP <- "Event locations"
EVENT_POINT_RADIUS <- 4.5
EVENT_COLOR_2026 <- "#00FFFF"
EVENT_COLOR_2025 <- "#FF1493"
EVENT_COLOR_OTHER <- "#000000"

ROR_BREAKPOINTS <- c(0.4, 0.7, 1.5, 5, 15)
CHANGE_RATIO_BREAKPOINTS <- c(0.5, 0.7, 1.3, 1.5, 2)
ROR_VALUE_COLORS <- c("green4", "lightgreen", "white", "red", "darkred", "purple4")
CHANGE_RATIO_VALUE_COLORS <- c("blue3", "lightblue", "white", "orange1", "red2", "red4")
TOP_PERCENTILE_COLOR <- "red"

raster_cache <- new.env(parent = emptyenv())

event_report_products <- list(
  ror = list(
    label = "Relative odds ratio",
    years = PREDICTION_YEARS,
    year_label = function(year) as.character(year),
    raster_path = function(year) {
      file.path(PREDICTION_DIR, "ror", sprintf("ror_%s.tif", year))
    },
    value_layer = "ROR_mean",
    top_layer = "ROR_mean_top1pct",
    breakpoints = ROR_BREAKPOINTS,
    colors = ROR_VALUE_COLORS,
    legend_title = "Mean ROR",
    top_legend_title = "Top 1% ROR"
  ),
  ror_anomaly = list(
    label = "Historical ROR anomaly ratio",
    years = ANOMALY_YEARS,
    year_label = function(year) as.character(year),
    raster_path = function(year) {
      file.path(PREDICTION_DIR, "hist_anom", sprintf("hist_anom_%s.tif", year))
    },
    value_layer = "histROR_anomaly_mean",
    top_layer = "histROR_anomaly_mean_top1pct",
    breakpoints = CHANGE_RATIO_BREAKPOINTS,
    colors = CHANGE_RATIO_VALUE_COLORS,
    legend_title = "Mean ratio",
    top_legend_title = "Top 1% anomaly ratios"
  )
)

available_years <- function(product) {
  product$years[vapply(product$years, function(year) file.exists(product$raster_path(year)), logical(1))]
}

missing_products <- names(event_report_products)[
  vapply(event_report_products, function(product) length(available_years(product)) == 0, logical(1))
]
if (length(missing_products) > 0) {
  stop(
    "No raster files were found for report product(s): ",
    paste(missing_products, collapse = ", "),
    ". Run 08_train_predict_full_superlearner_historical_anomaly.R first.",
    call. = FALSE
  )
}

read_report_raster <- function(product_id, year) {
  cache_key <- paste(product_id, year, sep = "_")
  if (exists(cache_key, envir = raster_cache, inherits = FALSE)) {
    return(get(cache_key, envir = raster_cache))
  }
  product <- event_report_products[[product_id]]
  path <- product$raster_path(year)
  if (!file.exists(path)) {
    stop("Missing raster for report: ", path, call. = FALSE)
  }
  raster <- terra::rast(path)
  assign(cache_key, raster, envir = raster_cache)
  raster
}

read_report_layer <- function(product_id, year, layer_type) {
  product <- event_report_products[[product_id]]
  layer_name <- if (identical(layer_type, "top")) product$top_layer else product$value_layer
  raster <- read_report_raster(product_id, year)
  if (!layer_name %in% names(raster)) {
    stop("Raster does not contain layer ", layer_name, ": ", product$raster_path(year), call. = FALSE)
  }
  raster[[layer_name]]
}

layer_values <- function(raster_layer) {
  values <- terra::values(raster_layer, mat = FALSE)
  values[is.finite(values)]
}

product_value_bins <- function(product_id) {
  values <- unlist(
    lapply(available_years(event_report_products[[product_id]]), function(year) {
      layer_values(read_report_layer(product_id, year, "value"))
    }),
    use.names = FALSE
  )
  if (length(values) == 0) {
    stop("No finite raster values were found for ", event_report_products[[product_id]]$label, call. = FALSE)
  }
  breaks <- event_report_products[[product_id]]$breakpoints
  lower <- min(values, na.rm = TRUE)
  upper <- max(values, na.rm = TRUE)
  unique(c(lower, breaks[breaks > lower & breaks < upper], upper))
}

discrete_break_labels <- function(bins) {
  labels <- character(length(bins) - 1)
  for (i in seq_along(labels)) {
    lower <- bins[i]
    upper <- bins[i + 1]
    if (i == length(labels)) {
      labels[i] <- paste0(signif(lower, 3), "+")
    } else {
      labels[i] <- paste0(signif(lower, 3), "-", signif(upper, 3))
    }
  }
  labels
}

fit_product_bounds <- function(map, product_id) {
  product <- event_report_products[[product_id]]
  years <- available_years(product)
  extents <- lapply(years, function(year) terra::ext(read_report_raster(product_id, year)))
  xmin <- min(vapply(extents, terra::xmin, numeric(1)), na.rm = TRUE)
  xmax <- max(vapply(extents, terra::xmax, numeric(1)), na.rm = TRUE)
  ymin <- min(vapply(extents, terra::ymin, numeric(1)), na.rm = TRUE)
  ymax <- max(vapply(extents, terra::ymax, numeric(1)), na.rm = TRUE)
  leaflet::fitBounds(map, lng1 = xmin, lat1 = ymin, lng2 = xmax, lat2 = ymax)
}

standardize_event_overlay_columns <- function(events) {
  names(events) <- tolower(names(events))
  if (!"longitude" %in% names(events) && "long" %in% names(events)) {
    events$longitude <- events$long
  }
  if (!"latitude" %in% names(events) && "lat" %in% names(events)) {
    events$latitude <- events$lat
  }
  if (!"id" %in% names(events)) {
    events$id <- seq_len(nrow(events))
  }
  if (!"type" %in% names(events)) {
    events$type <- NA_character_
  }
  if (!"country" %in% names(events)) {
    events$country <- NA_character_
  }
  events$year <- as.integer(events$year)
  events$latitude <- as.numeric(events$latitude)
  events$longitude <- as.numeric(events$longitude)
  events <- events[is.finite(events$latitude) & is.finite(events$longitude), , drop = FALSE]
  events$event_color <- ifelse(
    events$year == 2026,
    EVENT_COLOR_2026,
    ifelse(events$year == 2025, EVENT_COLOR_2025, EVENT_COLOR_OTHER)
  )
  events$popup_text <- sprintf(
    "<strong>%s</strong><br>Year: %s<br>Type: %s<br>Country: %s",
    htmltools::htmlEscape(events$id),
    events$year,
    htmltools::htmlEscape(ifelse(is.na(events$type) | !nzchar(events$type), "unknown", events$type)),
    htmltools::htmlEscape(ifelse(is.na(events$country) | !nzchar(events$country), "unknown", events$country))
  )
  events
}

read_event_overlay_data <- function() {
  outcomes_csv <- existing_path(OUTCOMES_CSV_CANDIDATES)
  if (is.na(outcomes_csv)) {
    warning("Could not find outcomes CSV for event overlay.")
    return(data.frame())
  }
  standardize_event_overlay_columns(utils::read.csv(outcomes_csv, stringsAsFactors = FALSE))
}

add_event_overlay <- function(map) {
  events <- read_event_overlay_data()
  if (nrow(events) == 0) {
    return(map)
  }
  map |>
    leaflet::addCircleMarkers(
      data = events,
      lng = ~longitude,
      lat = ~latitude,
      radius = EVENT_POINT_RADIUS,
      stroke = TRUE,
      weight = 1,
      color = ~event_color,
      fillColor = ~event_color,
      fillOpacity = 0.95,
      opacity = 1,
      popup = ~popup_text,
      label = ~id,
      group = EVENT_OVERLAY_GROUP
    ) |>
    leaflet::addLegend(
      position = "bottomright",
      colors = c(EVENT_COLOR_2026, EVENT_COLOR_2025, EVENT_COLOR_OTHER),
      labels = c("2026 events", "2025 events", "Earlier events"),
      title = "Events",
      opacity = 1
    )
}

enforce_single_active_overlay <- function(map, overlay_groups, default_group) {
  htmlwidgets::onRender(
    map,
    "
    function(el, x, data) {
      var map = this;
      var overlayGroups = data.overlayGroups || [];
      var defaultGroup = data.defaultGroup;
      var activeGroup = defaultGroup;
      var radioName = el.id + '-exclusive-year-overlay';
      function normalizeLabel(value) {
        return (value || '').replace(/\\s+/g, ' ').trim();
      }
      function makeOverlayControlsRadio() {
        var overlayPane = el.querySelector('.leaflet-control-layers-overlays');
        if (!overlayPane) return;
        overlayPane.querySelectorAll('label').forEach(function(label) {
          var input = label.querySelector('input');
          var groupName = normalizeLabel(label.textContent);
          if (!input || overlayGroups.indexOf(groupName) === -1) return;
          input.type = 'radio';
          input.name = radioName;
          input.checked = groupName === activeGroup;
        });
      }
      setTimeout(makeOverlayControlsRadio, 0);
      map.on('overlayadd', function(event) {
        if (overlayGroups.indexOf(event.name) !== -1) activeGroup = event.name;
        setTimeout(makeOverlayControlsRadio, 0);
      });
      map.on('overlayremove', function(event) {
        if (event.name === activeGroup) activeGroup = defaultGroup;
        setTimeout(makeOverlayControlsRadio, 0);
      });
    }
    ",
    data = list(overlayGroups = unname(overlay_groups), defaultGroup = unname(default_group))
  )
}

make_layered_report_map <- function(product_id, display_mode) {
  product <- event_report_products[[product_id]]
  years <- available_years(product)
  overlay_groups <- vapply(years, product$year_label, character(1))
  default_group <- overlay_groups[length(overlay_groups)]

  map <- leaflet::leaflet(options = leaflet::leafletOptions(preferCanvas = TRUE)) |>
    leaflet::addProviderTiles(leaflet::providers$Esri.WorldTopoMap, group = "Topo") |>
    leaflet::addProviderTiles(leaflet::providers$CartoDB.Positron, group = "Light") |>
    leaflet::addScaleBar(position = "bottomleft")

  if (identical(display_mode, "value")) {
    bins <- product_value_bins(product_id)
    legend_labels <- discrete_break_labels(bins)
    colors <- grDevices::colorRampPalette(product$colors)(length(legend_labels))
    finite_values <- unlist(
      lapply(years, function(year) layer_values(read_report_layer(product_id, year, "value"))),
      use.names = FALSE
    )
    palette <- leaflet::colorBin(
      palette = colors,
      bins = bins,
      domain = finite_values,
      na.color = LEAFLET_NA_COLOR
    )
    for (i in seq_along(years)) {
      map <- leaflet::addRasterImage(
        map,
        read_report_layer(product_id, years[i], "value"),
        colors = palette,
        opacity = MAP_OPACITY_VALUE,
        project = TRUE,
        maxBytes = 50 * 1024 * 1024,
        group = overlay_groups[i]
      )
    }
    map <- leaflet::addLegend(
      map,
      position = "bottomleft",
      colors = rev(colors),
      labels = rev(legend_labels),
      title = product$legend_title,
      opacity = MAP_OPACITY_VALUE
    )
  } else {
    palette <- leaflet::colorBin(
      palette = TOP_PERCENTILE_COLOR,
      bins = c(0.5, 1.5),
      domain = c(1),
      na.color = LEAFLET_NA_COLOR
    )
    for (i in seq_along(years)) {
      map <- leaflet::addRasterImage(
        map,
        read_report_layer(product_id, years[i], "top"),
        colors = palette,
        opacity = MAP_OPACITY_TOP1,
        project = TRUE,
        maxBytes = 50 * 1024 * 1024,
        group = overlay_groups[i]
      )
    }
    map <- leaflet::addLegend(
      map,
      position = "bottomleft",
      colors = TOP_PERCENTILE_COLOR,
      labels = "Top 1%",
      title = product$top_legend_title,
      opacity = MAP_OPACITY_TOP1
    )
  }

  map <- add_event_overlay(map)
  map |>
    leaflet::addLayersControl(
      baseGroups = c("Topo", "Light"),
      overlayGroups = c(overlay_groups, EVENT_OVERLAY_GROUP),
      options = leaflet::layersControlOptions(collapsed = FALSE)
    ) |>
    leaflet::hideGroup(setdiff(overlay_groups, default_group)) |>
    fit_product_bounds(product_id) |>
    enforce_single_active_overlay(overlay_groups, default_group)
}
```

```{css, echo=FALSE}
.main-container {
  max-width: 1800px !important;
  width: 98% !important;
}

.section-note {
  max-width: 1300px;
  font-size: 1.05rem;
  line-height: 1.45;
  margin-bottom: 1rem;
}

.leaflet {
  border: 1px solid #cccccc;
  margin-bottom: 1.5rem;
}
```

This self-contained HTML report reads the final full-data SuperLearner GeoTIFF rasters generated by `08_train_predict_full_superlearner_historical_anomaly.R` and embeds interactive Leaflet maps. Event points are shown as cyan for 2026, deeppink for 2025, and black for earlier years. Raster overlays are packaged into the knitted HTML file; basemap tiles still require an internet connection when the recipient opens the report.

## Overall Relative Odds Ratio Predictions

<div class="section-note">
These maps show the mean relative odds ratio (ROR) for each prediction-grid cell. A value above 1 means the estimated odds for that cell-year are higher than the study-area average odds for the same year.
</div>

### Mean ROR

```{r ror-value-map}
make_layered_report_map("ror", "value")
```

### Top 1% Mean ROR

```{r ror-top-map}
make_layered_report_map("ror", "top")
```

## Historical Anomaly In Relative Odds Ratios

<div class="section-note">
These maps show the ratio of the current year's mean ROR to the historical average mean ROR for the same cell, using all prediction years currently available in the final-model raster folder. Values above 1 indicate that the cell-year is elevated relative to its own historical baseline; values below 1 indicate that it is lower than its historical baseline.
</div>

### Mean Historical ROR Anomaly Ratio

```{r ror-anomaly-value-map}
make_layered_report_map("ror_anomaly", "value")
```

### Top 1% Mean Historical ROR Anomaly Ratios

```{r ror-anomaly-top-map}
make_layered_report_map("ror_anomaly", "top")
```
